Unknown

### 2025-06-20 ms/ms.qmd
working on ms

We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions). We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]. With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).

Table of directional predictions (with table summarizing results? part of conceptual figure? in supplement?)

caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats. For each hypothesis, we make predictions for how native light habitat, light treatment, and light intensity would affect AA.

|  hypothesis  | native light habitat | light treatment  | light intensity |
|:------------:|:--------------------:|:----------------:|:---------------:|
| acclimatory  | cor(PAR, AA) = 0   | AAsun = AAshade  | AA2000 \> AA150 |
|   plastic    | cor(PAR, AA) = 0 | AAsun \> AAshade | AA2000 = AA150  |
| constitutive | cor(PAR, AA) \> 0   | AAsun = AAshade  | AA2000 = AA150  |

@@

Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats. Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax. The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments. WORKING HERE

## Materials and Methods

>>>>

caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats. For each hypothesis, we make predictions for how native plant area index (PAI), growth light treatment, and measurement light intensity would affect AA.

|  hypothesis  | measurement light intensity | growth light intensity  | native PAI |
|:------------:|:---------------------------:|:------------------------:|:----------:|
| acclimatory  | AA2000 > AA150             | AAsun = AAshade          | cor(PAI, AA) = 0 |
| plastic      | AA2000 = AA150             | AAsun > AAshade          | cor(PAI, AA) = 0 |
| constitutive | AA2000 = AA150             | AAsun = AAshade          | cor(PAI, AA) > 0 |

<!-- Paragraph 5 (summarize methods) -->
We tested these hypotheses by comparing AA among amphistomatous wild tomato species (systematic cites) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions). We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]. With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details). We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods). The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation (GEDI; cite) and georeferenced accession collection information (cite TGRC). The growth light intensities were `r ppfd_sun` (sun treatment) and `r ppfd_shade` (shade treatment) while all other environment conditions were nearly identical (see @sec-methods for supporting detail). The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.

#### MARKER
(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum. After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down

@@

Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats. Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax. The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments. (not finished yet)

## Materials and Methods [#sec-methods]
